Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-25 Manuscript/manuscript.Rmd
edits

When evaluating model performances, we found that regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the other two predictor types (Figure 3). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations. Delving further, we found that when combined with SMOTE re-sampling, random forest classifiers built using distance-type predictors out performed other re-sampling techniques across resolutions, suggesting that a combination of both random over- and under-sampling can improve TAD boundary prediction. Interestingly, given these parameters, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. This reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the added sparcity in the interaction frequency matrices at finer resolutions. This sparcity corresponds to inaccurate TAD calling which leads to worse model performance. Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics (SMOTE resampling with distance type predictors). <!--These results imply that there is an ideal resolution in TAD boundary prediction, and that that an optimally performing cell line specific random forest classifier, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.--> Models implemented on the K562 cell line were found to yield the same optimal data level characteristics (Supplementary Figure 6).

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models, within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6 A). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS outperformed other cell line and annotation-specific models at all resolutions (Figure 6 B). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal.

<!--
![**The proportional importance of optimally built cell line specific models was outweighed by transcription factor binding sites across resolutions.** Stacked barplots illustrate the proportion of total predictive importance in cell line specific random forest classifiers for (A) GM12878 and (B) K562 respectively. Model inputs included the optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors). Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).](Figures/proportion_of_importances_v2.png)
-->

<!--
![**Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.](Figures/comparing_sets_of_annotations_v2.png)
-->

## PREDICTIVE IMPORTANCES CONFIRMED BIOLOGICAL FUNCTIONALITY OF KNOWN TRANSCRIPTION FACTOR BINDING SITES

@@

<!--
![**Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.](Figures/figure7_variable_importance_heatmaps_tfbs_gm12878.png)
-->

@@

<!--
## INFLUENTIAL TFBS TENDED TO BE CLOSER TO BINS THAT OVERLAPPED WITH FLANKED TAD BOUNDARIES THAN THOSE THAT DID NOT

We then investigated the spatial patterns of the most predictive TFBS around TAD boudaries by characterizing the distance profiles between the classes of genomic bins. For GM12878, at 10 kb resolution, we found that SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were all statistically significantly closer to bins that overlapped with flanked TAD boudaries than those that did not (Figure 8). [??? not there] Although still statistically significantly different, the distance profiles between genomic class were found to be much more similar at lower resolutions (Supplementary Figure 9). For the K562 cell line, we reported the distance profiles of the same transcription factor binding sites, excluding the YY1sc281, due to its unavailability in the UCSC Genome Browser. These cell line specific distance profiles were again found to be significantly different between genomic class, at all resolutions (Supplementary Figure 10).

![**Figure 8. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with 10 kb flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around 10 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].](Figures/figure8_distance_boxplots_key_tfbs_gm12878_10kb.png)
-->

>>>>

When evaluating model performances, we found that regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the other two predictor types (Figure 3). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations. Delving further, we found that when combined with SMOTE re-sampling, random forest classifiers built using distance-type predictors out performed other re-sampling techniques across resolutions, suggesting that a combination of both random over- and under-sampling can improve TAD boundary prediction. Interestingly, given these parameters, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. This reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the added sparcity in the interaction frequency matrices at finer resolutions. This sparcity corresponds to inaccurate TAD calling which leads to worse model performance. Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics (SMOTE resampling with distance type predictors). These results imply that there is an ideal resolution in TAD boundary prediction, and that that an optimally performing cell line specific random forest classifier, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling. Models implemented on the K562 cell line were found to yield the same optimal data level characteristics (Supplementary Figure 6).

@@

We then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (HM, ChromHMM, and TFBS). For optimal random forest classifiers, we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites (Supplementary Figure 7). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that models built simply on TFBS outperformed other annotation-specific models at all resolutions (Figure 4A; Supplementary Figure 8A). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 5 kb and 10 kb resolutions. Interestingly though, at the 25 kb, 50 kb, and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction.

##PREDICTIVE IMPORTANCES CONFIRMED BIOLOGICAL FUNCTIONALITY OF KNOWN ENRICHED TRANSCRIPTION FACTOR BINDING SITES AT TAD BOUNDARIES

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated, we sought to avoid the incidence of overfitting our optimally built classifier by performing recursive feature elimination (Supplementary Figure 9) to establish the necessary number of TFBS to include. We found that our cell line specific model was able to obtain near optimal performance using only ~8 TFBS (Figure 4B; Supplementary Figure 8B).

<!--
## PREDICTIVE IMPORTANCES CONFIRMED BIOLOGICAL FUNCTIONALITY OF KNOWN TRANSCRIPTION FACTOR BINDING SITES

@@

@@

-->
